---
title: "HW05"
author: "Robert Gruener"
date: "November 4, 2018"
output: github_document
---

### Let's Do Some Cancer Biology
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(broom)

theme_set(theme_bw())
```

I am in a cancer biology lab, and wanted to learn R to help with my research. So for this hw assignment, we are going to be using some data to see how cancer cell lines respond to various drug treatments. In particular, we are going to focus on triple-negative breast cancers (TNBC), a subtype of breast cancer that has poor prognosis and is currently treated with chemotherapies.

## Let's Get the Data

The data that I'll be using is originally from [GDSC](https://www.cancerrxgene.org/) and [CTRP](https://portals.broadinstitute.org/ctrp.v2.1/), but I placed files that were slightly modified already in the repo for this assignment. They are both large cancer cell line drug screens, but are independent of one another. 

The first cell line info page was from [cellosaurus](https://web.expasy.org/cellosaurus/), the second breast cancer cell line info is from [a recent review](http://www.jcancer.org/v08p3131.htm) and shows the molecular classifications of the breast cancer cell lines. These files are also in the repo. 

Shown below is some of the information these data files contain. 

```{r, data_download, cache=TRUE}
print("GDSC")
(GDSC <- read_tsv(file = "GDSC_2016_Drug_Data.txt"))

print("CTRP")
(CTRP <- read_tsv(file = "CTRPv2_AUC.txt"))

print("Cell Line Info")
(cell_line_info <- readxl::read_xlsx(path = "All_Cell_Line_Info_10-30-17.xlsx"))

print("Breast cancer cell line info")
(breast_cell_line_info <- read_csv(file = "Breast_CCL_Subtypes.csv"))
```

## Cleaning and Subsetting the data

We now need to combine the info files with the drug data, and subset to just the breast cancer cell lines

```{r, clean_subset}
GDSC_breast <- GDSC %>% 
  select(Drug.name, Drug.Id, Cell.line.name, Tissue, Avg.IC50) %>% #selecting relevant columns
  filter(Tissue== "breast") %>% #filter to breast cancers only
  #need to remove - from cell line names to make matches easier, chose to remove all punctuation as follows:
  mutate(CCL_Name = gsub("[[:punct:]]", replacement = "", x = Cell.line.name)) %>% 
  #need to join with breast cancer subtype data
  left_join(., breast_cell_line_info, by = c("CCL_Name" = "Cell lines"))


#I need to subset CTRP to breast cancers as well, but that information is not in here yet
#so I will join it with the cell line information

CTRP_breast <- CTRP %>% 
  left_join(., cell_line_info, by= c("CCL_Name" = "Cell_Line_Name")) %>% 
  select(CCL_Name, Tissue, Cellosaurus_Id, cpd_name, Avg_AUC) %>% 
  filter(Tissue == "breast") %>% 
  #need to remove - from cell line names to make matches easier, chose to remove all punctuation
  mutate(CCL_Name = gsub("[[:punct:]]", replacement = "", x = CCL_Name)) %>% 
  left_join(., breast_cell_line_info, by = c("CCL_Name" = "Cell lines"))
```

Here is how the data looks once cleaned  
``` {r, echo = TRUE}
GDSC_breast
CTRP_breast
``` 


Unfortunately, there is likely some missing subtype information still. Let's see:

``` {r, clean_subset_missing}
# Unfortunately, there is likely some missing subtype information still. Let's see:

missing_gdsc <- GDSC_breast %>% 
    select(CCL_Name, Subtype) %>% 
    filter(is.na(Subtype)) %>% 
    unique()

missing_CTRP <- CTRP_breast %>% 
    select(CCL_Name, Subtype) %>% 
    filter(is.na(Subtype)) %>% 
    unique()

either_missing <- bind_rows(missing_CTRP, missing_gdsc) %>% 
    unique()
knitr::kable(either_missing)

```

Those 12 values will be manually curated. The top 10 rows of the final cleaned up data frame without any missing NA values and with the relevant columns selected is displayed below for GDSC and CTRP respectively: 

```{r, clean_subset_2}
#Yup, there are 12 missing values. Hang-on while I go manually curate them...
# these sites helped with the manual curation, 
# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3127435/, https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5001206/

missing_subtype <- c("TNB", "H", "LA", "UN", "TNB", "UN", "TN", "LB", "TN", "H", "UN", "TNB") #"UN", unclassifiable, problematic cell lines with conflicting status claims in the literature
names(missing_subtype) <- either_missing$CCL_Name
either_missing$Subtype <- missing_subtype

#now, to replace those NA values with the manually curated values above
GDSC_breast <- GDSC_breast %>% 
  mutate(Subtype= if_else(is.na(Subtype), missing_subtype[CCL_Name], Subtype)) %>% 
  select(cpd_name = Drug.name, CCL_Name, Subtype, Avg_IC50 = Avg.IC50)

CTRP_breast <- CTRP_breast %>% 
  mutate(Subtype= if_else(is.na(Subtype), missing_subtype[CCL_Name], Subtype)) %>%
  select(cpd_name, CCL_Name, Subtype, Avg_AUC) %>% 
  unique()

#now let's see how the data looks:
GDSC_breast
CTRP_breast

```

## Number of Cell Lines per Subtype

Now that things look tidy, lets count the number of cell lines in each breast cancer group. 

```{r, count, echo= FALSE}
knitr::kable(GDSC_breast %>% 
  select(CCL_Name, Subtype) %>%
  unique() %>% 
  count(Subtype), caption = "GDSC", format = "markdown")

knitr::kable(CTRP_breast %>% 
  select(CCL_Name, Subtype) %>%
  unique() %>% 
  count(Subtype), caption = "CTRP", format = "markdown")

```

Side-note, if anyone knows why the caption feature on kable doesn't seem to print anything, let me know. 

## Comparing Cell Line Response Across BC subtypes

Because of the low n-values, we will compare mean response between TNBC cells to non-TNBC cells. 

```{r, compare_response}

#make a new column to seprate TNBC from Other
GDSC_breast <- GDSC_breast %>% 
  mutate(TNBC = ifelse(Subtype %in% c("TN", "TNA", "TNB"), "yes", "no"))

CTRP_breast <- CTRP_breast %>% 
  mutate(TNBC = ifelse(Subtype %in% c("TN", "TNA", "TNB"), "yes", "no"))

GDSC_breast
CTRP_breast
``` 

With that, now we will get into the statistics. Below are the significant results for both datasets:

```{r, compare_results_statistics}
GDSC_t.test_results <- GDSC_breast %>% 
  group_by(cpd_name) %>% 
  do(tidy(t.test(Avg_IC50 ~ TNBC, data = .))) %>% # thanks second answer here
  # https://stackoverflow.com/questions/26244321/dplyr-summarise-multiple-columns-using-t-test
  ##I now need to change names because it is unclear what the values are. 
  #it goes in alphabetical order, so Estimate is the mean IC50 when TNBC=="no" minus the mean of TNBC=="yes" 
  select(cpd_name, Effect_notTNBC.minus.TNBC = estimate, p.value) %>% 
  ungroup() %>% #this is necessary inorder to adjust the p.values
  mutate(FDR = p.adjust(p.value, method = "fdr"), 
         more_sensitive = if_else(Effect_notTNBC.minus.TNBC > 0, "TNBC", "non-TNBC")) #this summarizes 
          ##the effect direction. The smaller the number, the more sensitive the cell was to this drug
          ###So if the values are positive it means TNBC was more negative and thus more sensitive

# Lets do the same thing for the other dataset

CTRP_t.test_results <- CTRP_breast %>% 
  #this next chunk is necessary for CTRP, because there are some cell lines that were only tested against one subtype, so a t-test isn't possible:
  group_by(cpd_name) %>% 
  filter(length(Avg_AUC)>9) %>% #after trial and error, 
  ## 9 is the minimum number of unique observations that doesn't throw an error. 
  ### I think this is fine, since having less than 9 total cell lines for a compound is unlikely to give sig. results anyway. 
  do(tidy(t.test(data = ., Avg_AUC~TNBC))) %>% # thanks second answer here
  # https://stackoverflow.com/questions/26244321/dplyr-summarise-multiple-columns-using-t-test
  ##I now need to change names because it is unclear what the values are. 
  #it goes in alphabetical order, so Estimate is the mean IC50 when TNBC=="no" minus the mean of TNBC=="yes" 
  select(cpd_name, Effect_notTNBC.minus.TNBC = estimate, p.value) %>% 
  ungroup() %>% #this is necessary inorder to adjust the p.values
  mutate(FDR = p.adjust(p.value, method = "fdr"), 
         more_sensitive = if_else(Effect_notTNBC.minus.TNBC > 0, "TNBC", "non-TNBC")) #this summarizes 
          ##the effect direction. The smaller the number, the more sensitive the cell was to this drug
          ###So if the values are positive it means TNBC was more negative and thus more sensitive

#Let's Print The Most Significant Results:
print("GDSC")
knitr::kable(GDSC_t.test_results %>% filter(FDR < 0.05) %>% arrange(FDR))
print("CTRP")  
knitr::kable(CTRP_t.test_results %>% filter(FDR < 0.05) %>% arrange(FDR))


```

As we can see, there are `r length(GDSC_t.test_results$FDR[GDSC_t.test_results$FDR < 0.05])` significant results in GDSC after multiple comparision correction. For CTRP, the number is only `r length(CTRP_t.test_results$FDR[CTRP_t.test_results$FDR < 0.05])`

Also, it is interesting to note that typically the direction is in favor of TNBC being *more sensitive*. In GDSC, even if we are very liberal and use a p.value of 0.1 (note, p-value and not FDR), there are `r sum(GDSC_t.test_results[GDSC_t.test_results$p.value < 0.10,]$more_sensitive == "TNBC")` drugs that are more effective on TNBC cells and only `r sum(GDSC_t.test_results[GDSC_t.test_results$p.value < 0.10,]$more_sensitive == "non-TNBC")`. 
For CTRP, there are `r sum(CTRP_t.test_results[CTRP_t.test_results$p.value < 0.10,]$more_sensitive == "TNBC")` compounds that generally show more sensitivity in TNBC, and only `r sum(CTRP_t.test_results[CTRP_t.test_results$p.value < 0.10,]$more_sensitive == "non-TNBC")` that have the opposite effect direction. 
This is surprising because TNBC is generally considered more resistant to therapy in comparison to the other breast cancer subtypes.


## Graphing the result for select drugs

I now would like to graph some key results for some drugs I am interested in. Since I am going to make the same kind of graph for 6 different drugs, I'm going to first make a graphing_function

### Graphing Function
I'm choosing to show the code for the graphing function, just because there isn't much tables, etc. I could do with it. 

```{r, graphing_function, echo = TRUE}

drug_grapher <- function(drug, dataset){
  if(dataset == "GDSC_breast") { #must be the appropriate dataset
    drug_plot <- GDSC_breast %>%   
      filter(cpd_name == drug) %>% #filters to the drug provided
      ggplot(aes(x=TNBC, y= Avg_IC50, fill = TNBC)) + 
        geom_boxplot() + #makes box plot based on TNBC status and drug response
        geom_point() + #adds points to visualize the individual values
        labs(x = "TNBC status",
            y = "Average IC50", 
            title = paste(drug, "Response in TNBC vs non-TNBC CCLs", sep = " "), 
            caption = dataset) + 
        annotate("text", -Inf, Inf,  #adds the p-values and FDR values as text to the graph 
                 label = paste("FDR of", GDSC_t.test_results$FDR[GDSC_t.test_results$cpd_name == drug], sep = " "), 
                 hjust = 0, vjust = 3) +
      annotate("text", -Inf, Inf,  
               label = paste("P-value of", GDSC_t.test_results$p.value[GDSC_t.test_results$cpd_name == drug], sep = " "), 
               hjust = 0, vjust = 1) +
        theme(legend.position = "none")
    return(drug_plot)
  } #close first if statement 
  if(dataset == "CTRP_breast") { #same thing for the other dataset
    drug_plot <- CTRP_breast %>% 
      filter(cpd_name == drug) %>% 
      ggplot(aes(x=TNBC, y= Avg_AUC, fill = TNBC)) +
      geom_boxplot() +
      geom_point() + 
      labs(x = "TNBC status",
           y = "Average AUC", 
           title = paste(drug, "Response in TNBC vs non-TNBC CCLs", sep = " "), 
           caption = dataset) + 
      annotate("text", -Inf, Inf,  
               label = paste("FDR of", CTRP_t.test_results$FDR[CTRP_t.test_results$cpd_name == drug], sep = " "), 
               hjust = 0, vjust = 3) +
      annotate("text", -Inf, Inf,  
               label = paste("P-value of", CTRP_t.test_results$p.value[CTRP_t.test_results$cpd_name == drug], sep = " "), 
               hjust = 0, vjust = 1) + 
      theme(legend.position = "none")
    return(drug_plot)
  } #close if
} #close function

```


### Graphing results

Now, finally, I'll be graphing the results of those 6 select drugs. 

```{r, graphs}

drugs_of_interest <- list()  #making a list so I can pair the drug and dataset vectors  
drugs_of_interest[[1]] <- c("VAF-347", "I-BET151", "JQ-1", "GSK525762A", "I-BET-762", "PFI-1", "MK-1775", "birinapant")
drugs_of_interest[[2]] <- c("CTRP_breast", "CTRP_breast", "CTRP_breast", "CTRP_breast", "GDSC_breast", "GDSC_breast", "CTRP_breast", "CTRP_breast") 

#now runs a for loop to graph all the data
#NB: with ggplot to get the plots to print to the console in a for loop, you must use the print function
for (i in 1:length(drugs_of_interest[[1]])) {
  print(drug_grapher(drug = drugs_of_interest[[1]][i], dataset = drugs_of_interest[[2]][i]))
} 

```

As we can see from the graphs, of the drugs we are interested in ("VAF-347", "I-BET151", "JQ-1", "GSK525762A", "I-BET-762", "PFI-1", "MK-1775", "birinapant"), only the drug MK-1775 shows a significant increase in sensitivity in the TNBC cell lines. That is, based on these cell line results, MK-1775 looks to be more effective in TNBC compared to other breast cancer subtypes. 



## Session Info:

```{r, echo= TRUE}
sessionInfo()
```


